---
title: "Introduction to linear modelling"
author: "BIOSCI220"
date: "Semester 1 2020"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: yeti
    toc_depth: 3
    highlight: espresso
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = TRUE,message = FALSE, warning = FALSE)
library(dplyr)
library(ggplot2)
library(magick)
library(gganimate)
```

## Learing Objectives

By the end of this lab student should be able to

   + Develop a biologically relevant question of interest from the dataset and identify the following types of variables in the dataset
      + response variable
      + explanatory variable
   + Express their question of interest accurately and concisely
   + Carry out and interpret tests for the existence of relationships between explanatory variables and the response in a linear model
   + Write R code to fit a linear model with a single continuous explanatory variable
   + Write R code to fit a linear model with a continuous explanatory variable and a factor explanatory variable
   + Interpret estimated effects with reference to confidence intervals from linear regression models. Specifically the interpretation of
      + the intercept
      + the effect of a factor
      + the effect of a one-unit increase in a numeric variable
      + the effect of an x-unit increase in a numeric variable
   + MAke a point prediction of the response for a new observation

### Some great `r icon::fa_twitter(colour = "#1DA1F2")` threads

 <blockquote class="twitter-tweet"><p lang="en" dir="ltr">I thought it could be helpful to have a thread on ANOVA in R. As a statistical consultant, this is the most frequent FAQ I get from clients - how to run a linear model on their data, conduct hypothesis tests, extract predicted means and perform contrasts.</p>&mdash; We are R-Ladies (@WeAreRLadies) <a href="https://twitter.com/WeAreRLadies/status/1223790298024726528?ref_src=twsrc%5Etfw">February 2, 2020</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

 
 <blockquote class="twitter-tweet"><p lang="en" dir="ltr">I&#39;ve made this cheat sheet and I think it&#39;s important. Most stats 101 tests are simple linear models - including &quot;non-parametric&quot; tests. It&#39;s so simple we should only teach regression. Avoid confusing students with a zoo of named tests. <a href="https://t.co/9PFR1ly3lW">https://t.co/9PFR1ly3lW</a> 1/n</p>&mdash; Jonas K. Lindeløv (@jonaslindeloev) <a href="https://twitter.com/jonaslindeloev/status/1110907133833502721?ref_src=twsrc%5Etfw">March 27, 2019</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>
 
 <blockquote class="twitter-tweet"><p lang="en" dir="ltr">The aov() function in <a href="https://twitter.com/hashtag/Rstats?src=hash&amp;ref_src=twsrc%5Etfw">#Rstats</a> is actually a wrapper around the lm() function <a href="https://t.co/FbvxQdtD4c">pic.twitter.com/FbvxQdtD4c</a></p>&mdash; Dan Quintana (@dsquintana) <a href="https://twitter.com/dsquintana/status/1189536788790042625?ref_src=twsrc%5Etfw">October 30, 2019</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>
 
 
## Line of best fit

<iframe src="https://script.google.com/macros/s/AKfycbw2qx1b8iTZZXY5-aaaaGp76XiutxS1iuCFmL24IyBz6GACuSML/exec"width="600" height="350"/>

## (Most) Hypothesis tests are just linear regression

```{r regression gif, fig.cap= "Gif inspired by [https://crumplab.github.io/statistics/gifs.html](https://crumplab.github.io/statistics/gifs.html)"}
d <- mtcars
fit <- lm(mpg ~ hp, data = d)
d$predicted <- predict(fit)   ## predicted values
d$residuals <- residuals(fit) ## residuals

coefs<-coef(lm(mpg ~ hp, data = mtcars))

x<-d$hp
move_line<-c(seq(-6,6,.5),seq(6,-6,-.5))
total_error<-length(length(move_line))
cnt<-0
for(i in move_line){
  cnt<-cnt+1
  predicted_y <- coefs[2]*x + coefs[1]+i
  error_y <- (predicted_y-d$mpg)^2
  total_error[cnt]<-sqrt(sum(error_y)/32)
}

move_line_sims<-rep(move_line,each=32)
total_error_sims<-rep(total_error,each=32)
sims<-rep(1:50,each=32)

d<-d %>% slice(rep(row_number(), 50))

d<-cbind(d,sims,move_line_sims,total_error_sims)


anim<-ggplot(d, aes(x = hp, y = mpg, frame=sims)) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgrey",lty = 2) +  
  geom_abline(intercept = 30.09886+move_line_sims, slope = -0.06822828)+
  lims(x = c(0,400), y = c(-10,40))+
  geom_segment(aes(xend = hp, yend = predicted+move_line_sims, color="red"), alpha = .5) + 
  geom_point() +
  theme_classic()+
  theme(legend.position="none")+
  xlab("Explanatory variable")+ylab("Response variable")+
  transition_manual(frames=sims)+
  enter_fade() + 
  exit_fade()+
  ease_aes('sine-in-out')

animate(anim,fps=5)

```


## Example and `R` code

[Carbon footprint data](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-02-18/readme.md)

[San Fransisco trees](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-01-28/readme.md)

[NZ bird of the year](https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-11-19)

[Global PLastic waste](https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-05-21)